---
title: "Synthetic data"
subtitle: "What is simulated, how, and what the masking preserves"
---

```{r}
#| label: setup
source("R/setup.R")
library(survival)
library(knitr)
```

## Why synthetic data, and what survives the masking

The original analysis used real patient records, so none of it can be published. Simply
renaming the columns of a real dataset is not a solution — the values, the counts, and
the joint distribution would all still be real. The alternative taken here is to
regenerate the data from scratch, preserving the properties that make the analysis
non-trivial and discarding everything else.

What is preserved:

- **Roles.** Each column plays the same part it played originally: identifier, exposure,
  demographic, comorbidity, lab measure, event indicator, follow-up time, competing
  event.
- **Confounding.** The same covariates drive treatment assignment and the outcome. This
  is the whole reason a propensity score is needed; without it, balance adjustment would
  be decoration.
- **Correlation between covariates.** The six binary clinical variables form a chain,
  and the continuous measure depends on age and on two of the binaries. Independent
  covariates would make every balance table look good by accident.
- **A variable on the causal pathway.** `clinical_continuous_1` is affected by the
  comorbidities and in turn affects the outcome. It is deliberately excluded from the
  propensity model and entered last in the Cox ladder.
- **Competing risks with month-precision death dates**, in Study B.
- **Two real data defects**: a literal `"null"` string in a character column, and a
  small number of negative follow-up times.

What is discarded: every real value, every real count, the true cohort size, the
therapeutic area, the outcome definitions, and the study period.

What is *gained*: the true parameter values are known. Every notebook that follows
checks its estimates against them.

## The schema

```{r}
#| label: schema
data.frame(
  variable = c(ID_VAR, GROUP_VAR, DEMOGRAPHIC_VARS, CLINICAL_BINARY_VARS,
               CLINICAL_CONTINUOUS_VAR),
  role = c(
    "Identifier",
    "Exposure (reference level listed first)",
    "Demographic — continuous", "Demographic — binary",
    "Demographic — categorical, contains a missingness code",
    rep("Baseline comorbidity — binary", length(CLINICAL_BINARY_VARS)),
    "Baseline lab measure — continuous, on the causal pathway"
  ),
  in_propensity_model = c(
    "No", "Treatment", rep("Yes", length(DEMOGRAPHIC_VARS)),
    rep("Yes", length(CLINICAL_BINARY_VARS)), "No — see below"
  )
) |> kable()
```

`clinical_continuous_1` is left out of the propensity model on purpose. It is influenced
by the comorbidities and influences the outcome, so conditioning on it would adjust away
part of the very effect being estimated. Its exclusion is also what makes it the
covariate that fails balance first — which is what the caliper search has to solve.

## The data-generating process

Survival times come from a Weibull proportional hazards model. The survivor function is

$$S(t \mid x) = \exp\left\{-\left(\tfrac{t}{\lambda}\right)^{k}\exp(x'\beta)\right\}$$

so that inverting at $U \sim \mathrm{Uniform}(0,1)$ gives event times directly, and
$\exp(\beta)$ is a hazard ratio in the ordinary Cox sense. Censoring times are drawn
separately, and follow-up is the minimum of the event time, the censoring time and the
administrative horizon.

The ordering matters. A common shortcut is to draw a follow-up time first and then draw
an event indicator from a model that uses that time as a predictor. That produces data
where no hazard ratio exists to be estimated, so every downstream Cox model is
estimating a parameter the data does not have.

```{r}
#| label: dgp-code
#| code-fold: true
#| eval: false
# From R/simulate.R
weibull_ph_time <- function(lp, shape, scale0) {
  u <- runif(length(lp))
  scale0 * (-log(u) * exp(-lp))^(1 / shape)
}

t_evt  <- weibull_ph_time(lp, WEIBULL_SHAPE, scale0)
t_cens <- rexp(n, rate = cens_rate)
t_obs  <- pmin(t_evt, t_cens, STUDY_A_ADMIN_YEARS)
event  <- as.integer(t_evt <= pmin(t_cens, STUDY_A_ADMIN_YEARS))
```

### True parameters

```{r}
#| label: true-params
true_tbl <- data.frame(
  quantity = c("Study A, outcome 1 — treatment",
               "Study A, outcome 2 — treatment",
               "Study B, event of interest — treatment",
               "Study B, competing death — treatment"),
  true_hr = round(exp(c(TRUE_A_OUTCOME_1[["treat"]], TRUE_A_OUTCOME_2[["treat"]],
                        TRUE_B_EVENT[["treat"]], TRUE_B_DEATH[["treat"]])), 3)
)
kable(true_tbl)
```

Treatment lowers the hazard of the Study B event of interest but is essentially null on
the competing death hazard. That asymmetry is deliberate: it is what makes the
cause-specific and subdistribution hazard ratios differ in a direction known in advance,
and therefore what makes fitting both of them informative rather than redundant.

`demo_race` appears in **no** outcome model. Its true effect is exactly zero, which makes
it the yardstick for reading the random forest importance rankings later.

## Generating the cohorts

```{r}
#| label: generate
study_a <- load_study("a", refresh = TRUE)
study_b <- load_study("b", refresh = TRUE)

data.frame(
  study = c("A", "B"),
  n = c(nrow(study_a), nrow(study_b)),
  arms = c(paste(levels(study_a$exposure_group), collapse = " / "),
           paste(levels(study_b$exposure_group), collapse = " / "))
) |> kable()
```

## Face validity

A simulation is only useful if it produces what it claims to produce. These checks are
assertions, not printouts: if the generator drifts, the notebook fails to render rather
than quietly publishing a wrong result.

### Study A: event rates and follow-up

```{r}
#| label: check-a-rates
a_valid <- study_a[study_a$followup_time_1 > 0 & study_a$followup_time_2 > 0, ]

rates_a <- data.frame(
  endpoint = c("Outcome 1", "Outcome 2"),
  events = c(sum(a_valid$outcome_1), sum(a_valid$outcome_2)),
  event_rate = round(c(mean(a_valid$outcome_1), mean(a_valid$outcome_2)), 4),
  median_followup = round(c(median(a_valid$followup_time_1),
                            median(a_valid$followup_time_2)), 2)
)
kable(rates_a)

stopifnot(rates_a$events > 500)
```

### Study A: the exposure is genuinely confounded

If treatment were assigned at random, the unadjusted and adjusted hazard ratios would
agree and there would be nothing for a propensity score to do.

```{r}
#| label: check-a-confounding
adj_formula <- reformulate(
  c(GROUP_VAR, "I((demo_age - 65)/10)", "demo_sex",
    "clinical_binary_1", "clinical_binary_2", "clinical_binary_4",
    "clinical_binary_6", "I((clinical_continuous_1 - 80)/10)"),
  response = "Surv(followup_time_1, outcome_1)"
)

fit_unadj <- coxph(Surv(followup_time_1, outcome_1) ~ exposure_group, data = a_valid)
fit_adj   <- coxph(adj_formula, data = a_valid)

true_hr_1 <- exp(TRUE_A_OUTCOME_1[["treat"]])

data.frame(
  model = c("Unadjusted", "Fully adjusted", "Truth"),
  hr = round(c(exp(coef(fit_unadj))[1], exp(coef(fit_adj))[1], true_hr_1), 3)
) |> kable()
```

The unadjusted estimate sits close to the null while the true hazard ratio is clearly
protective. Confounding here does not merely bias the estimate — it hides the effect
entirely. That is the situation the rest of Study A exists to handle.

```{r}
#| label: assert-a
# The adjusted model must land near the truth; the unadjusted one must not.
stopifnot(
  abs(log(exp(coef(fit_adj))[1]) - log(true_hr_1)) < 0.10,
  exp(coef(fit_unadj))[1] > exp(coef(fit_adj))[1] + 0.10
)
```

### Study B: competing-risk structure

```{r}
#| label: check-b
b_prepared <- prepare_outcome_data(study_b)

status_tbl <- as.data.frame(table(b_prepared$cr_status))
names(status_tbl) <- c("status", "n")
status_tbl$percent <- round(100 * status_tbl$n / nrow(b_prepared), 2)
kable(status_tbl)
```

```{r}
#| label: check-b-hr
cs_formula <- reformulate(
  c(GROUP_VAR, "I((demo_age - 65)/10)", "demo_sex", "clinical_binary_1",
    "clinical_binary_3", "I((clinical_continuous_1 - 80)/10)"),
  response = "Surv(fu_years, cs_event)"
)
fit_cs <- coxph(cs_formula, data = b_prepared)

data.frame(
  model = c("Cause-specific, unadjusted", "Cause-specific, adjusted", "Truth"),
  hr = round(c(
    exp(coef(coxph(Surv(fu_years, cs_event) ~ exposure_group, data = b_prepared)))[1],
    exp(coef(fit_cs))[1],
    exp(TRUE_B_EVENT[["treat"]])
  ), 3)
) |> kable()

stopifnot(abs(log(exp(coef(fit_cs))[1]) - TRUE_B_EVENT[["treat"]]) < 0.12)
```

### The injected defects

```{r}
#| label: check-defects
data.frame(
  defect = c("Race coded as the literal string \"null\"",
             "Negative follow-up time, outcome 1",
             "Negative follow-up time, outcome 2",
             "Death date rounded back before index date (Study B)"),
  n = c(sum(study_a$demo_race == "null"),
        sum(study_a$followup_time_1 < 0),
        sum(study_a$followup_time_2 < 0),
        attr(b_prepared, "n_death_dates_clamped"))
) |> kable()

stopifnot(sum(study_a$demo_race == "null") > 0,
          sum(study_a$followup_time_1 < 0) > 0)
```

Each of these forces a decision that a clean dataset would let you skip.

`"null"` is a string, not `NA`, so it survives a naive read and only announces itself
when a factor gains an unexpected level. Treating it as a category, dropping it, or
recoding it to missing are all defensible — silently coercing it to `NA` via
`as.numeric()` is not, because it is indistinguishable from data that was never
collected.

Negative follow-up times cannot be right and have to be removed before modelling, with
the count reported rather than absorbed.

The clamped death dates are a consequence of month precision: a death recorded in the
same month as enrolment rounds back to the first of that month, which can fall before
the index date. A death cannot precede enrolment, so the date is clamped forward and the
count is reported. This is the first sign of an issue that
[the competing-risks notebook](05-competing-risks.qmd) returns to.

## What gets cached

```{r}
#| label: cache
data.frame(
  file = c(PATHS$study_a, PATHS$study_b),
  rows = c(nrow(study_a), nrow(study_b)),
  size_kb = round(c(file.size(PATHS$study_a), file.size(PATHS$study_b)) / 1024, 1)
) |> kable()
```

Both files are derived from a fixed seed and are not tracked in version control.
Deleting `data/synthetic/` and re-rendering reproduces them exactly.
