---
title: "Balance and propensity score matching"
subtitle: "Problem 1 — matching ran, but it did not balance"
---

```{r}
#| label: setup
source("R/setup.R")
library(MatchIt)
library(tableone)
library(cobalt)
library(knitr)

study_a <- require_schema(
  clean_study_a(load_study("a")),
  schema_for("study_a_clean"),
  stage = "Study A balance and matching"
)
```

## The problem

The first propensity score match ran without error, returned a matched cohort, and
produced a balance table that looked broadly reasonable. Several covariates were still
imbalanced past the conventional threshold.

The failure mode is worth naming precisely, because it is silent: `matchit()` will
always return matched pairs. Nearest-neighbour matching finds the closest available
control for each treated subject whether or not that control is *close*. Without a
caliper, "closest available" can be arbitrarily far away. Nothing errors, nothing warns,
and the output has the shape you expected.

So the check has to be explicit, and it has to happen every time.

## Attrition

```{r}
#| label: attrition
kable(attr(study_a, "attrition"))
```

Race is kept as an explicit `Missing` category rather than dropped. Dropping those
subjects would assume the missingness is unrelated to everything else; keeping it as a
level makes no such assumption and lets the balance table show whether the two arms
differ in how often it occurs.

## How balance is judged

The absolute standardized mean difference, with 0.10 as the threshold and 0.05 as
comfortable. Not a hypothesis test.

At `r format(nrow(study_a), big.mark = ",")` subjects, a t-test has enough power to
flag differences far too small to bias anything. A significant imbalance and a
meaningful imbalance are different claims, and only the second one matters here. The SMD
is also unitless and unaffected by sample size, so the same threshold means the same
thing in the full cohort and in a subgroup.

One implementation note that costs people time: in `tableone`, `smd` is an argument to
`print()`, not to `CreateTableOne()`. Passing it to the latter silently does nothing.

```{r}
#| label: table1
tab1 <- CreateTableOne(
  vars = BASELINE_PREDICTORS, strata = GROUP_VAR,
  data = study_a, test = FALSE, addOverall = TRUE
)
print(tab1, smd = TRUE, showAllLevels = TRUE, printToggle = FALSE) |>
  as.data.frame() |> kable()
```

## Before matching

```{r}
#| label: pre-balance
pre_balance <- balance_table(study_a, GROUP_VAR, BASELINE_PREDICTORS)
pre_balance[order(-pre_balance$abs_smd), ] |> kable(row.names = FALSE, digits = 4)
```

```{r}
#| label: pre-summary
n_pre_imbalanced <- sum(!pre_balance$balanced)
```

`r n_pre_imbalanced` covariate rows exceed the threshold before matching, led by
`demo_age` and `clinical_binary_1` — the covariates that drive treatment assignment most
strongly.

`clinical_continuous_1` is also imbalanced, and it is a special case. It is not in the
propensity model, because it sits on the causal pathway: conditioning on it would adjust
away part of the effect being estimated. But it still has to *balance*, and matching can
only move it indirectly, through its correlation with the covariates that are in the
model. A covariate you cannot match on but must still balance is the one most likely to
decide whether a design works.

## Attempt 1 — nearest neighbour, no caliper

```{r}
#| label: match-naive
set.seed(PROJECT_SEED)
m_naive <- matchit(
  reformulate(PS_FORMULA_VARS, response = GROUP_VAR),
  data = study_a, method = "nearest", distance = "glm", ratio = 1
)
md_naive <- match.data(m_naive)

denoms <- balance_denominators(study_a, GROUP_VAR, BASELINE_PREDICTORS)
bal_naive <- balance_table(md_naive, GROUP_VAR, BASELINE_PREDICTORS, denoms = denoms)
bal_naive[order(-bal_naive$abs_smd), ] |> kable(row.names = FALSE, digits = 4)
```

```{r}
#| label: naive-verdict
data.frame(
  quantity = c("Matched subjects", "Percent of cohort retained",
               "Covariate rows still imbalanced",
               "Largest absolute SMD — before matching",
               "Largest absolute SMD — after matching"),
  value = c(format(nrow(md_naive), big.mark = ","),
            paste0(round(100 * nrow(md_naive) / nrow(study_a), 1), "%"),
            sum(!bal_naive$balanced),
            round(max(pre_balance$abs_smd), 4),
            round(max(bal_naive$abs_smd), 4))
) |> kable()
```

Matching without a caliper made balance **worse**. The largest absolute SMD rose from
`r round(max(pre_balance$abs_smd), 3)` before matching to
`r round(max(bal_naive$abs_smd), 3)` after it, and `demo_age` — the strongest driver of
treatment assignment — is further apart in the matched cohort than it was in the full
one.

That is not a paradox. Almost nobody was discarded
(`r round(100 * nrow(md_naive) / nrow(study_a), 1)`% retained), so every treated subject
was paired with whatever control was nearest, however far away that was. Nearest-neighbour
matching without a caliper is greedy: it works through the treated subjects in order, and
by the end the only controls left are the poor ones. The result is a cohort that has been
reshaped without being balanced.

This is the failure the notebook is named for. Nothing errored. The object has the right
class, the table has the right shape, and the sample size looks generous. The only way to
find out is to compute the balance table and read it.

```{r}
#| label: assert-naive-worse
# Lock the finding: an uncalipered match is not merely inadequate here, it is
# actively worse than doing nothing.
stopifnot(max(bal_naive$abs_smd) > max(pre_balance$abs_smd))
```

One detail that decides whether that table is trustworthy: the SMDs here use the
*pre-match* pooled standard deviations (`denoms`). Recomputing the denominator inside the
matched cohort is a quiet way to make balance look better than it is, because matching
narrows the covariate distributions and a smaller denominator shrinks every SMD. Before
and after have to be on one scale to be comparable.

## Attempt 2 — a fixed caliper

The standard next move is a caliper on the propensity score, expressed in standard
deviations of the linear predictor.

```{r}
#| label: match-fixed
set.seed(PROJECT_SEED)
m_fixed <- matchit(
  reformulate(PS_FORMULA_VARS, response = GROUP_VAR),
  data = study_a, method = "nearest", distance = "glm", ratio = 1,
  caliper = 0.20, std.caliper = TRUE
)
md_fixed <- match.data(m_fixed)
bal_fixed <- balance_table(md_fixed, GROUP_VAR, BASELINE_PREDICTORS, denoms = denoms)

data.frame(
  quantity = c("Matched subjects", "Percent of cohort retained",
               "Covariates still imbalanced", "Largest absolute SMD"),
  value = c(format(nrow(md_fixed), big.mark = ","),
            paste0(round(100 * nrow(md_fixed) / nrow(study_a), 1), "%"),
            sum(!bal_fixed$balanced),
            round(max(bal_fixed$abs_smd), 4))
) |> kable()
```

```{r}
#| label: fixed-worst
bal_fixed[order(-bal_fixed$abs_smd), ][1:5, ] |>
  kable(row.names = FALSE, digits = 4)
```

Better, and in the original work this is roughly where a first pass stopped. Whether it
is *good enough* is a question the table has to answer, not the fact that the code ran.

## Attempt 3 — search the caliper instead of guessing it

Rather than trying values by hand, walk a descending grid and take the **loosest**
caliper that balances every covariate. Direction matters: tighter calipers buy balance
by discarding subjects, so the loosest one meeting the threshold is the one that keeps
the most data and the most precision.

Two properties of the search are deliberate. Every attempt is logged, including the
failures — the log is the evidence that balance was verified rather than assumed. And if
no caliper in the grid succeeds, the function raises an error instead of returning its
best failure, because a matched cohort that never achieved balance is worse than no
matched cohort at all.

```{r}
#| label: caliper-search
search <- search_caliper(
  study_a, treat_var = GROUP_VAR,
  ps_vars = PS_FORMULA_VARS, balance_vars = BASELINE_PREDICTORS
)
kable(search$attempts, row.names = FALSE)
```

```{r}
#| label: caliper-selected
selected  <- search$selected
md_final  <- selected$matched
bal_final <- selected$balance

save_table(search$attempts, "psm_caliper_attempts")

data.frame(
  quantity = c("Selected caliper", "Matched subjects",
               "Percent of cohort retained", "Largest absolute SMD"),
  value = c(selected$caliper, format(nrow(md_final), big.mark = ","),
            paste0(round(100 * nrow(md_final) / nrow(study_a), 1), "%"),
            round(max(bal_final$abs_smd), 4))
) |> kable()
```

```{r}
#| label: assert-balance
# The whole point of the notebook: assert both that the selected caliper
# balances, and that a looser one failed. If nothing failed, the search was
# not demonstrating anything.
stopifnot(
  all(bal_final$balanced),
  max(bal_final$abs_smd) <= SMD_THRESHOLD,
  any(!search$attempts$balanced)
)
```

### The trade-off, stated plainly

```{r}
#| label: tradeoff
#| fig-height: 4
library(ggplot2)

ggplot(search$attempts, aes(x = caliper)) +
  geom_line(aes(y = max_abs_smd), linewidth = 0.7) +
  geom_point(aes(y = max_abs_smd, colour = balanced), size = 2.5) +
  geom_hline(yintercept = SMD_THRESHOLD, linetype = "dashed") +
  scale_x_reverse() +
  scale_colour_manual(values = c(`FALSE` = "#c0392b", `TRUE` = "#27ae60")) +
  labs(
    x = "Caliper (standard deviations, tightening to the right)",
    y = "Largest absolute SMD",
    colour = "All covariates balanced",
    title = "Balance improves as the caliper tightens",
    subtitle = "Dashed line is the 0.10 threshold"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Every step to the right also discards subjects. The `n_matched` column in the attempt
log is the price, and reporting it is part of reporting the match — a balanced cohort
that kept a third of the data is a different result from a balanced cohort that kept
nearly all of it.

## Balance before and after

```{r}
#| label: love-plot
#| fig-height: 6
love.plot(
  m_naive, stats = "mean.diffs", abs = TRUE,
  thresholds = c(m = SMD_THRESHOLD), var.order = "unadjusted",
  line = TRUE, drop.distance = TRUE,
  sample.names = c("Unmatched", "No caliper"),
  title = "Covariate balance, no caliper"
)
```

```{r}
#| label: love-plot-final
#| fig-height: 6
love.plot(
  selected$fit, stats = "mean.diffs", abs = TRUE,
  thresholds = c(m = SMD_THRESHOLD), var.order = "unadjusted",
  line = TRUE, drop.distance = TRUE,
  sample.names = c("Unmatched", paste0("Caliper ", selected$caliper)),
  title = paste0("Covariate balance, caliper ", selected$caliper)
)
```

## Matched Table 1

```{r}
#| label: table1-matched
tab1_matched <- CreateTableOne(
  vars = BASELINE_PREDICTORS, strata = GROUP_VAR,
  data = md_final, test = FALSE
)
print(tab1_matched, smd = TRUE, showAllLevels = TRUE, printToggle = FALSE) |>
  as.data.frame() |> kable()

save_table(bal_final, "psm_balance_final")
```

## The age subgroup

The subgroup gets its own match. Subsetting an already-matched cohort down to
`demo_age >= 65` would carry over pairs formed under the full-cohort propensity score,
and a score that balances the whole cohort does not balance a subgroup of it — the
covariate distributions inside the subgroup are different, so the score is fitted on the
wrong population.

```{r}
#| label: subgroup
study_a_65 <- age_subgroup(study_a)

search_65 <- search_caliper(
  study_a_65, treat_var = GROUP_VAR,
  ps_vars = PS_FORMULA_VARS, balance_vars = BASELINE_PREDICTORS
)
kable(search_65$attempts, row.names = FALSE)
```

```{r}
#| label: subgroup-result
md_65 <- search_65$selected$matched
save_table(search_65$attempts, "psm_caliper_attempts_age65")

data.frame(
  cohort = c("Full cohort", "Age >= 65"),
  n_eligible = c(nrow(study_a), nrow(study_a_65)),
  selected_caliper = c(selected$caliper, search_65$selected$caliper),
  n_matched = c(nrow(md_final), nrow(md_65)),
  max_abs_smd = round(c(max(bal_final$abs_smd),
                        max(search_65$selected$balance$abs_smd)), 4)
) |> kable()

stopifnot(all(search_65$selected$balance$balanced))
```

The two cohorts select different calipers. That is the expected result, not a bug — and
it is why a caliper carried over from another analysis is a value that happens to have
worked somewhere else.

## Saving the matched cohorts

```{r}
#| label: save
saveRDS(list(full = md_final, age65 = md_65,
             caliper_full = selected$caliper,
             caliper_65 = search_65$selected$caliper),
        file.path(dirname(PATHS$study_a), "study_a_matched.rds"))
```

## What to carry forward

- Matching that runs is not matching that worked. The only evidence of balance is a
  balance table computed after the fact, on pre-match denominators.
- A caliper is a tuning parameter, not a constant. Search it, log every attempt
  including the failures, and report the sample size each one costs.
- Fail loudly when no setting achieves balance. Returning the best available failure
  hands a downstream analysis a cohort that quietly does not support it.
- Subgroups need their own propensity score and their own caliper.

Next: [the Cox models](02-cox-models.qmd) that use this matched cohort.
