---
title: "Cox models"
subtitle: "Study A outcome models on the matched cohort"
---

```{r}
#| label: setup
source("R/setup.R")
library(survival)
library(survminer)
library(knitr)

study_a <- clean_study_a(load_study("a"))
matched <- readRDS(file.path(dirname(PATHS$study_a), "study_a_matched.rds"))
md_full <- matched$full
md_65   <- matched$age65
```

This notebook takes the matched cohort from
[the previous step](01-balance-and-psm.qmd) and estimates the treatment effect. Because
the data is simulated, each estimate can be compared against the hazard ratio that
generated it — which is the check the original analysis had no way to perform.

## The adjustment ladder

Covariates are added in blocks rather than all at once, so that the movement of the
treatment estimate is visible:

| Model | Adds | Rationale |
|---|---|---|
| m1 | Treatment only | Unadjusted contrast within the matched cohort |
| m2 | Demographics | Age, sex, race |
| m3 | Comorbidities | The six binary clinical variables |
| m4 | Lab measure | `clinical_continuous_1`, entered last |

The lab measure comes last on purpose. It is affected by the comorbidities and in turn
affects the outcome, so it sits on the causal pathway rather than being a pure
confounder. Adjusting for it removes part of the effect being estimated. Entering it in
its own step makes that visible rather than burying it in a single fully-adjusted model.

Matching is not a substitute for adjustment. A matched cohort is balanced on average, not
exactly, and residual imbalance within the caliper still biases the estimate. Fitting the
ladder on the matched data — matching *and* adjusting — is the standard recommendation.

```{r}
#| label: ladder-spec
ladder <- list(
  m1 = character(0),
  m2 = DEMOGRAPHIC_VARS,
  m3 = c(DEMOGRAPHIC_VARS, CLINICAL_BINARY_VARS),
  m4 = c(DEMOGRAPHIC_VARS, CLINICAL_BINARY_VARS, CLINICAL_CONTINUOUS_VAR)
)
ladder_labels <- c(m1 = "Treatment only", m2 = "+ demographics",
                   m3 = "+ comorbidities", m4 = "+ lab measure")
```

## Fitting

Matched pairs are not independent observations, so every model carries
`cluster(subclass)`. Without it the standard errors treat two members of a pair as two
independent subjects, and the confidence intervals come out too narrow.

```{r}
#| label: fit-ladder
fit_ladder <- function(data, spec, time_var, event_var, cohort_label,
                       outcome_label) {
  rows <- list()
  for (nm in names(spec)) {
    rhs <- c(GROUP_VAR, spec[[nm]], "cluster(subclass)")
    fml <- reformulate(rhs, response = sprintf("Surv(%s, %s)", time_var, event_var))
    fit <- coxph(fml, data = data)

    term <- paste0(GROUP_VAR, STUDY_A_GROUP_LEVELS[2])
    rows[[nm]] <- model_result_row(
      fit, term = term, label = cohort_label, model = ladder_labels[[nm]],
      se_source = "robust",
      extra = list(outcome = outcome_label,
                   n = fit$n, events = fit$nevent)
    )
  }
  do.call(rbind, rows)
}

results <- list()
for (oc in names(STUDY_A_OUTCOMES)) {
  spec <- STUDY_A_OUTCOMES[[oc]]
  results[[paste0(oc, "_full")]] <- fit_ladder(
    md_full, ladder, spec$time_var, spec$event_var, "Full cohort", spec$label
  )
  results[[paste0(oc, "_65")]] <- fit_ladder(
    md_65, ladder, spec$time_var, spec$event_var, "Age >= 65", spec$label
  )
}
cox_results <- do.call(rbind, results)
rownames(cox_results) <- NULL
```

## Results

```{r}
#| label: results-table
true_hr <- c("Outcome 1" = exp(TRUE_A_OUTCOME_1[["treat"]]),
             "Outcome 2" = exp(TRUE_A_OUTCOME_2[["treat"]]))

display <- add_display(cox_results)
display$true_hr <- round(true_hr[display$outcome], 3)

display[, c("outcome", "analysis", "model", "n", "events", "hr_ci", "p", "true_hr")] |>
  kable(col.names = c("Outcome", "Cohort", "Model", "N", "Events",
                      "HR (95% CI)", "p", "True HR"),
        row.names = FALSE)

save_table(cox_results, "cox_results")
```

## Reading the ladder

```{r}
#| label: ladder-plot
#| fig-height: 5
library(ggplot2)

plot_df <- display[display$analysis == "Full cohort", ]
plot_df$model <- factor(plot_df$model, levels = ladder_labels)

ggplot(plot_df, aes(x = model, y = estimate, colour = outcome)) +
  geom_hline(yintercept = 1, colour = "grey60") +
  geom_pointrange(aes(ymin = ci_lower, ymax = ci_upper),
                  position = position_dodge(width = 0.35)) +
  geom_hline(aes(yintercept = true_hr, colour = outcome), linetype = "dashed") +
  labs(x = NULL, y = "Hazard ratio (95% CI)", colour = "Outcome",
       title = "Treatment effect across the adjustment ladder, full matched cohort",
       subtitle = "Dashed lines are the true hazard ratios that generated the data") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Did the design recover the truth?

The comparison that matters is not "unadjusted versus adjusted" — it is "unmatched versus
matched". Inside the matched cohort the confounding has already been dealt with, so `m1`
and `m3` should broadly agree. Outside it, they should not.

```{r}
#| label: recovery-check
final <- display[display$model == "+ comorbidities" &
                   display$analysis == "Full cohort", ]
matched_unadj <- display[display$model == "Treatment only" &
                           display$analysis == "Full cohort", ]

# The starting point: no matching, no adjustment.
raw <- study_a
unmatched_hr <- vapply(names(STUDY_A_OUTCOMES), function(oc) {
  s <- STUDY_A_OUTCOMES[[oc]]
  d <- raw[raw[[s$time_var]] > 0, ]
  exp(coef(coxph(reformulate(GROUP_VAR,
                             response = sprintf("Surv(%s, %s)", s$time_var, s$event_var)),
                 data = d)))[[1]]
}, numeric(1))

recovery <- data.frame(
  outcome = final$outcome,
  unmatched_unadjusted = round(unmatched_hr, 3),
  matched_unadjusted = round(matched_unadj$estimate, 3),
  matched_adjusted = round(final$estimate, 3),
  true_hr = final$true_hr,
  matched_adjusted_ci = final$hr_ci
)
kable(recovery, row.names = FALSE)
```

```{r}
#| label: recovery-distance
dist_tbl <- data.frame(
  outcome = final$outcome,
  d_unmatched = round(abs(log(unmatched_hr) - log(final$true_hr)), 3),
  d_matched_unadj = round(abs(log(matched_unadj$estimate) - log(final$true_hr)), 3),
  d_matched_adj = round(abs(log(final$estimate) - log(final$true_hr)), 3)
)
kable(dist_tbl, row.names = FALSE,
      col.names = c("Outcome", "Unmatched, unadjusted",
                    "Matched, unadjusted", "Matched, adjusted"))
```

Distance from the truth on the log-hazard scale. For outcome 1 the unmatched contrast is
off by `r dist_tbl$d_unmatched[1]`; matching alone cuts that to
`r dist_tbl$d_matched_unadj[1]`. Confounding was hiding the effect entirely, and the
match is what recovered it.

```{r}
#| label: assert-recovery
# The claim the design makes: matching moves the estimate substantially closer
# to the truth than doing nothing. Coverage of any single 95% interval is
# deliberately NOT asserted -- see below.
stopifnot(dist_tbl$d_matched_unadj < dist_tbl$d_unmatched,
          dist_tbl$d_matched_adj  < dist_tbl$d_unmatched)
```

Within the matched cohort, adding covariates barely moves outcome 1 (`m1`
`r recovery$matched_unadjusted[1]` → `m3` `r recovery$matched_adjusted[1]`). That
stability is itself a result: it is what balance looks like from the outcome side. If
adjustment had moved the estimate substantially, it would mean the match had left
confounding behind.

That said, adjustment still belongs in the model. Matching balances on average, not
exactly, and the residual imbalance inside the caliper is real. The original analysis went
back and forth on this — an early version fitted only the unadjusted model on matched
data, arguing the matching had already done the adjusting, and a later revision restored
the full ladder. The later position is the standard one.

Outcome 2's interval does not cover its true hazard ratio in this realization. That is
worth stating plainly rather than hiding, because it is the expected behaviour of a
confidence interval rather than a failure of the analysis: a 95% interval misses one time
in twenty, and this notebook produces four cohort-by-outcome combinations. Outcome 2 is
also the harder case — fewer events and a smaller true effect, so the least precise
estimate here.

The reason this can be diagnosed at all is that the truth is known. In the original
analysis an estimate landing 2.6 standard errors from its target would have been
indistinguishable from a coding error, and the only available response would have been to
re-read the code. Here the question is settled directly.

```{r}
#| label: precision-note
data.frame(
  outcome = final$outcome,
  events = final$events,
  se_log_hr = round(final$se_log, 4),
  sd_from_truth = round(abs(log(final$estimate) - log(final$true_hr)) / final$se_log, 2),
  covers_truth = final$ci_lower <= final$true_hr & final$ci_upper >= final$true_hr
) |> kable(row.names = FALSE)
```

Adding the lab measure in `m4` pulls both estimates toward the null. That is what
adjusting for a mediator does — it removes the part of the treatment effect that operates
through that variable. The `m4` estimate is not a better-adjusted version of `m3`; it is
an estimate of a different quantity, the direct effect not mediated by the lab measure.
Reporting it as the headline result would understate the total effect.

## Survival curves

```{r}
#| label: km
#| fig-height: 6
km_fit <- survfit(Surv(followup_time_1, outcome_1) ~ exposure_group, data = md_full)

ggsurvplot(
  km_fit, data = md_full,
  risk.table = TRUE, conf.int = TRUE, pval = TRUE,
  xlab = "Years since index", ylab = "Event-free probability",
  legend.labs = STUDY_A_GROUP_LEVELS,
  palette = c("#2c3e50", "#c0392b"),
  ylim = c(0.6, 1),
  risk.table.height = 0.28,
  ggtheme = theme_minimal(base_size = 12)
)
```

## Proportional hazards

The Cox model assumes the hazard ratio is constant over follow-up. If it is not, the
reported hazard ratio is an average over a changing effect, and its value depends on how
long people were followed.

```{r}
#| label: ph-check
ph_rows <- list()
for (oc in names(STUDY_A_OUTCOMES)) {
  spec <- STUDY_A_OUTCOMES[[oc]]
  fml <- reformulate(
    c(GROUP_VAR, DEMOGRAPHIC_VARS, CLINICAL_BINARY_VARS),
    response = sprintf("Surv(%s, %s)", spec$time_var, spec$event_var)
  )
  z <- cox.zph(coxph(fml, data = md_full))
  ph_rows[[oc]] <- data.frame(
    outcome = spec$label,
    chisq = round(z$table["GLOBAL", "chisq"], 3),
    df = z$table["GLOBAL", "df"],
    p_value = z$table["GLOBAL", "p"],
    conclusion = ph_status_label(z$table["GLOBAL", "p"])
  )
}
ph_tbl <- do.call(rbind, ph_rows)
ph_tbl$p_value <- fmt_p(ph_tbl$p_value)
kable(ph_tbl, row.names = FALSE)

save_table(ph_tbl, "cox_ph_check")
```

The wording matters. A non-significant test is not evidence that the hazards are
proportional — it is an absence of evidence that they are not, which at this sample size
is a weaker statement than it sounds. The data here were generated from a Weibull
proportional hazards model, so proportionality genuinely holds; the test agreeing with
that is a check on the test, not a discovery.

```{r}
#| label: schoenfeld
#| fig-height: 4
zph <- cox.zph(coxph(
  Surv(followup_time_1, outcome_1) ~ exposure_group + demo_age + demo_sex,
  data = md_full
))
plot(zph[1], main = "Scaled Schoenfeld residuals — treatment, outcome 1")
abline(h = 0, lty = 2, col = "grey50")
```

## What to carry forward

- Match *and* adjust. Matching balances on average, not exactly, and the residual
  imbalance inside the caliper is still confounding.
- Matched pairs are not independent. `cluster(subclass)` is what keeps the standard
  errors honest.
- A variable on the causal pathway belongs in its own step, and the model that includes
  it answers a different question. Which one is the headline is a decision to state, not
  a default to inherit.
- "No evidence against proportional hazards" is the most a PH test can support.

Next: [random forest importance](03-random-forest.qmd), and why the matched and unmatched
models are not interchangeable.
