---
title: "Random forest importance"
subtitle: "Problem 2 — two models, two different questions"
---

```{r}
#| label: setup
source("R/setup.R")
library(caret)
library(ranger)
library(recipes)
library(themis)
library(pROC)
library(ggplot2)
library(knitr)

study_a <- clean_study_a(load_study("a"))
matched <- readRDS(file.path(dirname(PATHS$study_a), "study_a_matched.rds"))
md_full <- matched$full
```

## The problem

Two random forests were fitted during the original work and they disagreed about which
variables mattered. One was fitted on the unmatched cohort with the treatment variable
excluded; the other on the propensity-matched cohort with treatment included. When the
rankings came out different, the natural question was which one was right.

Neither. They answer different questions, and the disagreement is the expected result.

| | Model A | Model B |
|---|---|---|
| Data | Unmatched, full cohort | Propensity-matched cohort |
| Treatment variable | Excluded | Included |
| Question | Which baseline characteristics predict the outcome? | Does treatment carry predictive information once the arms are balanced? |

Model A is a baseline risk question — it describes who is at risk, in the population as
it actually is. Model B is a prediction question asked inside a balanced population,
where treatment can be compared without the covariate distributions differing between
arms.

Neither estimates a treatment effect. That point is the reason this notebook exists.

## Predictor sets

Leakage control here is a whitelist. The predictors are named explicitly and everything
else is left out by construction, rather than assembling all columns and subtracting a
list of forbidden ones. A blacklist has to anticipate every bad column; a whitelist only
has to name the good ones, and it fails safe when a new column appears.

```{r}
#| label: predictors
predictors_a <- BASELINE_PREDICTORS               # no treatment
predictors_b <- c(GROUP_VAR, BASELINE_PREDICTORS) # treatment included

data.frame(
  model = c("A — unmatched, no treatment", "B — matched, with treatment"),
  n_predictors = c(length(predictors_a), length(predictors_b)),
  predictors = c(paste(predictors_a, collapse = ", "),
                 paste(predictors_b, collapse = ", "))
) |> kable()
```

What is deliberately absent from both: the follow-up times, the other outcome, the
post-baseline indicator, the record identifier, and the `MatchIt` artifacts
(`distance`, `weights`, `subclass`) that appear in a matched data frame and would let the
model reconstruct the matching.

```{r}
#| label: assert-leakage
forbidden <- c(ID_VAR, "outcome_1", "outcome_2",
               "followup_time_1", "followup_time_2",
               "post_baseline_indicator", MATCHIT_ARTIFACT_VARS)

stopifnot(
  length(intersect(predictors_a, forbidden)) == 0,
  length(intersect(predictors_b, forbidden)) == 0,
  # The two sets differ by exactly the treatment variable.
  identical(setdiff(predictors_b, predictors_a), GROUP_VAR),
  length(setdiff(predictors_a, predictors_b)) == 0
)
```

`followup_time_*` deserves a specific mention. It is the most tempting column to leave in
— it is numeric, it is in the data, and it is enormously predictive. It is also
post-baseline: someone with a short follow-up time either had the event or left the study,
so including it lets the model read the answer off the input. A classification forest with
follow-up time among the predictors will report an excellent AUC and mean nothing.

## Fitting

```{r}
#| label: rf-fit
rf_model <- function(data, predictors, outcome_var, label, seed = PROJECT_SEED) {
  d <- data[, c(predictors, outcome_var)]
  d$.outcome <- factor(ifelse(d[[outcome_var]] == 1, "Event", "NoEvent"),
                       levels = c("NoEvent", "Event"))
  d[[outcome_var]] <- NULL

  set.seed(seed)
  idx   <- createDataPartition(d$.outcome, p = 0.75, list = FALSE)
  train <- d[idx, ]
  test  <- d[-idx, ]

  # The recipe is passed to caret::train rather than pre-baked. SMOTE creates
  # synthetic minority rows by interpolating between neighbours, so baking it
  # once before cross-validation puts rows derived from fold k into the
  # training data of every other fold. The tuning metric is then optimistic and
  # the selected hyperparameters are chosen against a leak.
  rec <- recipe(.outcome ~ ., data = train) |>
    step_dummy(all_nominal_predictors()) |>
    step_smote(.outcome, over_ratio = 0.5)

  ctrl <- trainControl(
    method = "cv", number = 5, classProbs = TRUE,
    summaryFunction = twoClassSummary, savePredictions = "final",
    allowParallel = FALSE
  )
  grid <- expand.grid(mtry = c(2, 5), splitrule = "gini",
                      min.node.size = c(10, 30))

  # Tuning runs without importance. Permutation importance is expensive and is
  # not used to select hyperparameters, so computing it on all 20 resample fits
  # would be pure waste -- it is computed once, on the final model.
  set.seed(seed)
  fit <- train(rec, data = train, method = "ranger", metric = "ROC",
               num.trees = RF_NUM_TREES, importance = "none",
               tuneGrid = grid, trControl = ctrl)

  # Refit the selected configuration once, with permutation importance. SMOTE
  # is applied to the whole training set here, which is correct: the leak this
  # avoids is synthetic rows crossing *resample* boundaries during tuning, and
  # tuning is already finished. step_smote carries skip = TRUE, so baking the
  # test set does not oversample it.
  prepped     <- prep(rec, training = train)
  train_baked <- bake(prepped, new_data = NULL)
  test_baked  <- bake(prepped, new_data = test)

  set.seed(seed)
  final_fit <- ranger(
    .outcome ~ ., data = train_baked, num.trees = RF_NUM_TREES,
    mtry = fit$bestTune$mtry, min.node.size = fit$bestTune$min.node.size,
    splitrule = "gini", importance = "permutation", probability = TRUE
  )

  prob <- predict(final_fit, data = test_baked)$predictions[, "Event"]
  roc_obj <- roc(response = test_baked$.outcome, predictor = prob,
                 levels = c("NoEvent", "Event"), direction = "<", quiet = TRUE)

  imp <- data.frame(variable = names(final_fit$variable.importance),
                    importance = as.numeric(final_fit$variable.importance),
                    row.names = NULL)
  imp <- imp[order(-imp$importance), ]
  imp$rank <- seq_len(nrow(imp))

  list(label = label, fit = final_fit, auc = as.numeric(auc(roc_obj)),
       importance = imp, n_train = nrow(train), n_test = nrow(test),
       best = fit$bestTune)
}
```

```{r}
#| label: run-models
model_a <- rf_model(study_a, predictors_a, "outcome_1",
                    "A — unmatched, no treatment")
model_b <- rf_model(md_full, predictors_b, "outcome_1",
                    "B — matched, with treatment")
```

```{r}
#| label: model-summary
data.frame(
  model = c(model_a$label, model_b$label),
  n_train = c(model_a$n_train, model_b$n_train),
  n_test = c(model_a$n_test, model_b$n_test),
  mtry = c(model_a$best$mtry, model_b$best$mtry),
  min_node = c(model_a$best$min.node.size, model_b$best$min.node.size),
  test_auc = round(c(model_a$auc, model_b$auc), 4)
) |> kable()

save_table(
  data.frame(model = c(model_a$label, model_b$label),
             auc = c(model_a$auc, model_b$auc)),
  "rf_auc_summary"
)
```

## The two rankings

```{r}
#| label: importance-tables
imp_a <- model_a$importance
imp_b <- model_b$importance
names(imp_a)[2:3] <- c("importance_A", "rank_A")
names(imp_b)[2:3] <- c("importance_B", "rank_B")

comparison <- merge(imp_a, imp_b, by = "variable", all = TRUE)
comparison <- comparison[order(comparison$rank_A), ]
kable(comparison, row.names = FALSE, digits = 2)

save_table(comparison, "rf_importance_comparison")
```

```{r}
#| label: importance-plot
#| fig-height: 6
plot_df <- rbind(
  transform(model_a$importance[1:10, ], model = "A — unmatched, no treatment"),
  transform(model_b$importance[1:10, ], model = "B — matched, with treatment")
)

ggplot(plot_df, aes(x = reorder(variable, importance), y = importance)) +
  geom_col(fill = "#2c3e50") +
  coord_flip() +
  facet_wrap(~model, scales = "free_y") +
  labs(x = NULL, y = "Permutation importance",
       title = "Variable importance depends on the question being asked") +
  theme_minimal(base_size = 11)
```

## Why they differ

Three reasons, all legitimate:

**The analytic samples are different.** Matching discards subjects outside the caliper
and reshapes the covariate distributions of those that remain. A forest splits on
whatever separates the data in front of it, so a different sample produces different
splits — before any question of which variables are "really" important.

**The predictor sets are different.** Once treatment is available, the model no longer
has to lean on variables that correlate with it. Importance is not a fixed property of a
variable; it is a property of a variable *in the presence of the others*. Adding one
predictor can demote several.

**Treatment acts as a proxy.** In observational data the treatment variable carries
information about prescribing patterns, clinical profile, and unmeasured severity — not
just the pharmacological effect. A model that finds treatment predictive has found that
treated and untreated subjects differ, which is what confounding is.

## What a high importance rank does not mean

Here the claim can be checked rather than argued, because the true effects are known.

```{r}
#| label: truth-vs-importance
truth <- data.frame(
  variable = c(GROUP_VAR, "demo_age", "demo_sex", "clinical_binary_1",
               "clinical_binary_2", "clinical_binary_4", "clinical_binary_6",
               CLINICAL_CONTINUOUS_VAR, "demo_race"),
  true_log_hr = c(
    TRUE_A_OUTCOME_1[["treat"]],
    TRUE_A_OUTCOME_1[["demo_age_per_decade"]],
    TRUE_A_OUTCOME_1[["demo_sex"]],
    TRUE_A_OUTCOME_1[["clinical_binary_1"]],
    TRUE_A_OUTCOME_1[["clinical_binary_2"]],
    TRUE_A_OUTCOME_1[["clinical_binary_4"]],
    TRUE_A_OUTCOME_1[["clinical_binary_6"]],
    TRUE_A_OUTCOME_1[["clinical_cont_per10"]],
    0
  )
)

imp_b_base <- model_b$importance
imp_b_base$base <- sub("_(Group_B|Missing|X1|1)$", "", imp_b_base$variable)
agg <- aggregate(importance ~ base, data = imp_b_base, FUN = max)
names(agg) <- c("variable", "max_importance")

check <- merge(truth, agg, by = "variable", all.x = TRUE)
check$true_effect_size <- round(abs(check$true_log_hr), 3)
check <- check[order(-check$max_importance), ]
kable(check[, c("variable", "max_importance", "true_log_hr", "true_effect_size")],
      row.names = FALSE, digits = 3)
```

```{r}
#| label: headline-ranks
race_rank <- imp_a$rank_A[grepl("^demo_race_Group_B", imp_a$variable)][1]
age_rank  <- imp_a$rank_A[imp_a$variable == "demo_age"][1]
```

Two results stand out, and both are checkable only because the truth is known.

**A variable with no effect at all ranks third.** `demo_race` appears in no outcome model
in the data-generating process — its true effect is exactly zero — yet
`demo_race_Group_B` comes `r race_rank`th of `r nrow(imp_a)` in the unmatched model, above
most of the comorbidities that genuinely drive the outcome. Permutation importance
measures how much a variable contributes to *predictions*, and a high-variance binary with
no causal role can still support locally useful splits. Non-zero importance is not
evidence of an effect, and a respectable rank is not either.

**The variable with the largest true effect ranks `r age_rank`th.** `demo_age` has the
biggest coefficient in the outcome model
(`r round(TRUE_A_OUTCOME_1[["demo_age_per_decade"]], 2)` per decade), and it lands near the
bottom. The reason is redundancy: age drives the comorbidities in this data, so by the
time the forest has split on `clinical_binary_1` and `clinical_binary_2` it already knows
most of what age would have told it. Permuting age then costs little accuracy, because the
information survives elsewhere. Importance is diluted across correlated predictors, and a
low rank can mean "this variable is redundant" rather than "this variable does not matter".

Between them, these two rows are the whole argument. The ranking does not recover the
ordering of true effect sizes, in either direction: a null variable placed high and the
strongest variable placed low.

```{r}
#| label: rank-vs-truth
#| fig-height: 4
ggplot(check[!is.na(check$max_importance), ],
       aes(x = true_effect_size, y = max_importance)) +
  geom_point(size = 3, colour = "#2c3e50") +
  geom_text(aes(label = variable), hjust = -0.1, size = 3) +
  expand_limits(x = max(check$true_effect_size, na.rm = TRUE) * 1.6) +
  labs(x = "True effect size (|log hazard ratio|)",
       y = "Permutation importance",
       title = "Importance is not effect size",
       subtitle = "Model B — matched cohort, treatment included") +
  theme_minimal(base_size = 12)
```

Feature importance carries no sign, so it cannot say whether a variable raises or lowers
risk. It carries no confidence interval, so it cannot say whether a difference in rank is
distinguishable from noise. And it is not invariant to the other variables in the model.
None of that makes it useless — it is a reasonable description of what a model relies on —
but it is not a treatment effect and it does not become one by ranking highly.

The treatment effect for these data is in
[the Cox models](02-cox-models.qmd): a hazard ratio, with a direction, an interval, and a
population it refers to.

## What to carry forward

- Decide what question the forest is answering before choosing the data and the predictor
  set. "Unmatched without treatment" and "matched with treatment" are two designs, not a
  right and a wrong one.
- Use a whitelist for predictors. It fails safe; a blacklist fails silently.
- Keep post-baseline variables out, follow-up time most of all.
- Resample inside the cross-validation loop, not before it, or the tuning metric is
  measuring a leak.
- Report importance as what a model relies on. For a treatment effect, fit a model that
  estimates one.
