---
title: "Overlap weighting"
subtitle: "Study B — balancing without discarding anyone"
---

```{r}
#| label: setup
source("R/setup.R")
library(survey)
library(tableone)
library(cobalt)
library(ggplot2)
library(knitr)

study_b <- load_study("b")
study_b$demo_race <- factor(study_b$demo_race, levels = c("Group_A", "Group_B"))
study_b$demo_sex  <- factor(study_b$demo_sex, levels = c(0, 1))
for (v in CLINICAL_BINARY_VARS) {
  study_b[[v]] <- factor(study_b[[v]], levels = c(0, 1))
}
study_b$exposure_group <- factor(study_b$exposure_group, levels = STUDY_B_GROUP_LEVELS)

# Schema check at the seam. The weighting columns do not exist yet, so this
# validates the baseline block only -- the full study_b_weighted contract is
# checked where the next notebook receives this file.
require_schema(
  study_b,
  schema_for("study_b_weighted")[c(ID_VAR, GROUP_VAR, BASELINE_PREDICTORS,
                                   "baseline_date", "last_fu_date")],
  stage = "Study B weighting"
)
```

## Why the design changed

[Study A](01-balance-and-psm.qmd) achieved balance by tightening a caliper until every
covariate came in under 0.10 — and paid for it by discarding subjects. That is the
standard trade: a matched cohort is balanced because the subjects who could not be
matched are gone.

Weighting takes a different route. Instead of selecting subjects, it reweights them, so
everyone contributes and balance comes from the weights rather than from exclusion. This
notebook builds the three weighting schemes, compares what each one costs, and shows why
overlap weighting was the one carried forward.

## The propensity model

```{r}
#| label: ps-model
ps_formula <- reformulate(PS_FORMULA_VARS, response = GROUP_VAR)

ps_fit <- glm(ps_formula, data = study_b, family = binomial())
if (!ps_fit$converged) stop("Propensity model did not converge.")

study_b$ps <- predict(ps_fit, type = "response")
treat <- as.integer(study_b$exposure_group == STUDY_B_GROUP_LEVELS[2])

stopifnot(all(is.finite(study_b$ps)), all(study_b$ps > 0), all(study_b$ps < 1))

data.frame(
  quantity = c("N", "Treated", "Control", "PS minimum", "PS maximum"),
  value = c(format(nrow(study_b), big.mark = ","),
            sum(treat == 1), sum(treat == 0),
            round(min(study_b$ps), 4), round(max(study_b$ps), 4))
) |> kable()
```

Only baseline covariates enter the propensity model. Nothing measured after the index
date belongs here — conditioning on a post-baseline variable would adjust away part of
the treatment's effect, and no amount of balance on such a variable is worth having.

`r STUDY_B_GROUP_LEVELS[1]` and `r STUDY_B_GROUP_LEVELS[2]` are the two treatment arms.
`r STUDY_B_GROUP_LEVELS[1]` is the reference, so every hazard ratio in Study B reads as
"`r STUDY_B_GROUP_LEVELS[2]` relative to `r STUDY_B_GROUP_LEVELS[1]`", and the propensity
score is the modelled probability of receiving **`r STUDY_B_GROUP_LEVELS[2]`**.

### Reading the plot below

The propensity score collapses all `r length(PS_FORMULA_VARS)` baseline covariates into a
single number: given everything known about a subject at index, how likely were they to
receive `r STUDY_B_GROUP_LEVELS[2]`? A score of 0.8 describes someone who looks like a
typical `r STUDY_B_GROUP_LEVELS[2]` recipient; 0.15 describes someone who looks like a
typical `r STUDY_B_GROUP_LEVELS[1]` recipient. It says nothing about their outcome.

The plot draws one curve per arm, showing how that arm's subjects are distributed across
the score. The vertical axis is density rather than headcount — each curve integrates to
one on its own — so the shapes stay comparable even though the arms are different sizes.

There are two separate things to read off it, and they pull in opposite directions:

- **Horizontal separation between the curves is the confounding.** Under randomisation the
  two curves would lie on top of each other, because nothing measured at baseline would
  predict the assigned arm. Any daylight between them measures how far this study is from
  that ideal.
- **Vertical overlap is what licenses the adjustment.** Pick any point on the horizontal
  axis: if both colours are present there, subjects with that profile exist in both arms
  and can be compared. Where one colour drops to zero, the estimate for that region comes
  from extrapolating the model rather than from anybody real.

```{r}
#| label: ps-overlap
#| fig-height: 4
ggplot(study_b, aes(x = ps, fill = exposure_group)) +
  geom_density(alpha = 0.5, colour = NA) +
  scale_fill_manual(values = c("#2c3e50", "#c0392b")) +
  labs(x = sprintf("Propensity score: modelled P(%s | baseline covariates)",
                   STUDY_B_GROUP_LEVELS[2]),
       y = "Density", fill = "Arm actually received",
       title = "The arms differ at baseline, but their propensity scores overlap",
       subtitle = paste("Separation between the curves is the confounding.",
                        "The shared region is what makes adjustment possible.",
                        sep = "\n")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.subtitle = element_text(lineheight = 1.2))
```

```{r}
#| label: ps-by-arm
ps_summary <- do.call(rbind, lapply(levels(study_b$exposure_group), function(lv) {
  p <- study_b$ps[study_b$exposure_group == lv]
  data.frame(
    arm = lv,
    n = length(p),
    median_ps = round(median(p), 3),
    pct5 = round(quantile(p, 0.05), 3),
    pct95 = round(quantile(p, 0.95), 3)
  )
}))
kable(ps_summary, row.names = FALSE,
      col.names = c("Arm", "N", "Median PS", "5th pct", "95th pct"))
```

Two separate things are visible here, and they are easy to conflate.

**The curves are shifted apart, and that is the confounding.** Subjects who actually
received `r STUDY_B_GROUP_LEVELS[2]` have a higher modelled probability of receiving it —
median `r ps_summary$median_ps[2]` against `r ps_summary$median_ps[1]`. The propensity
model is not inventing that separation; it is measuring a real baseline difference between
the arms:

```{r}
#| label: why-they-differ
a_idx <- study_b$exposure_group == STUDY_B_GROUP_LEVELS[1]
c_idx <- study_b$exposure_group == STUDY_B_GROUP_LEVELS[2]
as_num <- function(x) as.numeric(as.character(x))

data.frame(
  covariate = c("Mean age", "clinical_binary_1 prevalence",
                "clinical_binary_2 prevalence"),
  arm_a = round(c(mean(study_b$demo_age[a_idx]),
                  mean(as_num(study_b$clinical_binary_1)[a_idx]),
                  mean(as_num(study_b$clinical_binary_2)[a_idx])), 3),
  arm_c = round(c(mean(study_b$demo_age[c_idx]),
                  mean(as_num(study_b$clinical_binary_1)[c_idx]),
                  mean(as_num(study_b$clinical_binary_2)[c_idx])), 3)
) |> kable(row.names = FALSE,
           col.names = c("Covariate", STUDY_B_GROUP_LEVELS[1], STUDY_B_GROUP_LEVELS[2]))
```

Older and sicker subjects were channelled toward `r STUDY_B_GROUP_LEVELS[2]`. Comparing
raw outcomes between the arms would credit the drug with a difference that baseline risk
created.

**The curves also overlap heavily, and that is what makes adjustment possible.** Across
roughly `r ps_summary$pct5[2]` to `r ps_summary$pct95[1]` both arms are well represented,
and `r format(sum(study_b$ps > 0.05 & study_b$ps < 0.95), big.mark = ",")` of
`r format(nrow(study_b), big.mark = ",")` subjects sit between 0.05 and 0.95.

This matters because every weighting scheme is answering a counterfactual: what would this
`r STUDY_B_GROUP_LEVELS[2]` subject's outcome have been under
`r STUDY_B_GROUP_LEVELS[1]`? A subject with a propensity score near 1 has almost no
counterpart in the other arm, so that question gets answered by extrapolation rather than
by evidence.

The picture to worry about is the opposite of this one: two curves that barely touch, or a
region where one arm's density is zero while the other's is high. No weighting scheme
repairs that — the response is to restrict the analysis to the region of common support.
Here the overlap is sound, so the design holds.

## Three sets of weights

```{r}
#| label: weights
w <- create_weights(study_b$ps, treat)
study_b <- cbind(study_b, w)

data.frame(
  method = c("IPTW", "Stabilized IPTW", "Overlap"),
  formula = c("1/ps for treated, 1/(1-ps) for control",
              "P(treated)/ps, P(control)/(1-ps)",
              "1-ps for treated, ps for control"),
  target_population = c("Whole population", "Whole population",
                        "Subjects with genuine clinical equipoise")
) |> kable()
```

The three differ in more than arithmetic. Inverse-probability weights target the whole
population and give the most weight to subjects who look least like their own arm — which
is exactly where the data is thinnest. A subject whose propensity score is far from their
observed treatment receives a very large weight and can move the estimate on their own:
the largest inverse-probability weight in this cohort is
`r round(max(w$iptw), 1)`, meaning one subject counts as much as
`r round(max(w$iptw))` others.

Overlap weights invert that. A subject with propensity score 0.5 — equally likely to have
received either treatment — gets the most weight, and subjects at the extremes get almost
none. Weights are bounded in (0, 1) by construction, so no individual can dominate.

The cost is that the estimand changes. Overlap weighting estimates the effect in the
population of subjects who could plausibly have received either treatment, not in the
whole population. That is a narrower question, and for a treatment comparison it is often
the more clinically meaningful one — but it has to be stated rather than assumed.

## What each weighting scheme costs

```{r}
#| label: diagnostics
diag <- weight_diagnostics(w, study_b$ps, treat)
kable(diag, row.names = FALSE)
save_table(diag, "weight_diagnostics")
```

```{r}
#| label: ess
ess_tbl <- data.frame(
  method = c("Unweighted", "IPTW", "Stabilized IPTW", "Overlap"),
  ess_total = round(c(nrow(study_b), diag$ess_total), 1)
)
ess_tbl$pct_of_cohort <- round(100 * ess_tbl$ess_total / nrow(study_b), 1)
kable(ess_tbl, row.names = FALSE)
```

The effective sample size is the honest accounting. Weighting does not create
information: a cohort of `r format(nrow(study_b), big.mark = ",")` subjects weighted so
that a few dominate carries the precision of a much smaller study. The Kish effective
sample size, $(\sum w)^2 / \sum w^2$, says how much smaller.

```{r}
#| label: weight-dist
#| fig-height: 4
long <- rbind(
  data.frame(method = "IPTW", weight = w$iptw),
  data.frame(method = "Stabilized IPTW", weight = w$sw_iptw),
  data.frame(method = "Overlap", weight = w$overlap_w)
)

ggplot(long, aes(x = weight)) +
  geom_histogram(bins = 60, fill = "#2c3e50") +
  facet_wrap(~method, scales = "free") +
  labs(x = "Weight", y = "Count",
       title = "Weight distributions",
       subtitle = "Note the differing x-axis scales") +
  theme_minimal(base_size = 11)
```

Note that the extreme-weight thresholds in the diagnostics table differ by method. A
single cutoff across all three would be meaningless: stabilized weights centre on 1,
untruncated inverse-probability weights have no upper bound, and overlap weights cannot
exceed 1 by construction. For overlap weighting the diagnostic reported instead is the
share of subjects whose propensity score falls outside the central range, since that is
where the design is actually under strain.

## Balance

```{r}
#| label: unweighted-table1
tab_unw <- CreateTableOne(vars = BASELINE_PREDICTORS, strata = GROUP_VAR,
                          data = study_b, test = FALSE, addOverall = TRUE)
print(tab_unw, smd = TRUE, showAllLevels = TRUE, printToggle = FALSE) |>
  as.data.frame() |> kable()
```

```{r}
#| label: weighted-table1
design <- svydesign(ids = ~1, data = study_b, weights = ~overlap_w)
tab_ow <- svyCreateTableOne(vars = BASELINE_PREDICTORS, strata = GROUP_VAR,
                            data = design, test = FALSE, addOverall = TRUE)
print(tab_ow, smd = TRUE, showAllLevels = TRUE, printToggle = FALSE) |>
  as.data.frame() |> kable()
```

`test = FALSE` is deliberate, for the same reason as in Study A: at this sample size a
hypothesis test detects differences far too small to matter, so the SMD column is the one
to read.

### All three methods side by side

```{r}
#| label: bal-tab
bal <- bal.tab(
  ps_formula, data = study_b, method = "weighting",
  weights = list(IPTW = w$iptw, `Stabilized IPTW` = w$sw_iptw,
                 Overlap = w$overlap_w),
  s.d.denom = "pooled", binary = "std", continuous = "std",
  thresholds = c(m = SMD_THRESHOLD), un = TRUE
)

# cobalt derives its column names from the names of the weights list and
# sanitises them, so the columns are located by pattern rather than assumed.
diff_cols <- grep("^Diff\\.", names(bal$Balance), value = TRUE)

bal_summary <- data.frame(
  method = sub("^Diff\\.", "", diff_cols),
  max_abs_smd = round(vapply(diff_cols, function(cl)
    max(abs(bal$Balance[[cl]]), na.rm = TRUE), numeric(1)), 6),
  row.names = NULL
)
bal_summary$method[bal_summary$method == "Un"] <- "Unweighted"
bal_summary$all_balanced <- bal_summary$max_abs_smd <= SMD_THRESHOLD
kable(bal_summary, row.names = FALSE)
save_table(bal_summary, "weighting_balance_summary")
```

```{r}
#| label: assert-balance
# Overlap weighting balances the covariates in its propensity model exactly, up
# to floating point. This is a mathematical property of the weights, not a
# result that needed tuning -- so it is asserted tightly.
stopifnot(bal_summary$max_abs_smd[bal_summary$method == "Overlap"] < 1e-6)
```

Overlap weighting achieves **exact** balance on every covariate in the propensity model —
not "under 0.10", but zero to numerical precision. This is a known algebraic property of
the weights rather than a lucky outcome: weighting the treated by $1-e$ and the controls
by $e$ makes the weighted covariate means identical by construction whenever the
propensity score comes from a logistic model containing those covariates.

That property is what makes the caliper search of Study A unnecessary here. There is no
tuning parameter, and no attempt log, because there is nothing to tune.

```{r}
#| label: love-plot
#| fig-height: 6
love.plot(
  bal, stats = "mean.diffs", abs = TRUE,
  thresholds = c(m = SMD_THRESHOLD), var.order = "unadjusted",
  line = TRUE, drop.distance = TRUE,
  title = "Covariate balance under three weighting schemes"
)
```

## The comparison that decided the design

```{r}
#| label: final-comparison
# Rebuild the Study A match if this notebook is rendered on its own, so it does
# not depend on another notebook having run first.
study_a_clean <- clean_study_a(load_study("a"))
matched_path  <- file.path(dirname(PATHS$study_a), "study_a_matched.rds")

matched_a <- if (file.exists(matched_path)) {
  readRDS(matched_path)
} else {
  s <- search_caliper(study_a_clean, GROUP_VAR, PS_FORMULA_VARS, BASELINE_PREDICTORS)
  list(full = s$selected$matched, caliper_full = s$selected$caliper)
}

data.frame(
  approach = c("Study A — 1:1 matching, tuned caliper",
               "Study B — overlap weighting"),
  subjects_used = c(
    sprintf("%s of %s (%.1f%%)",
            format(nrow(matched_a$full), big.mark = ","),
            format(nrow(study_a_clean), big.mark = ","),
            100 * nrow(matched_a$full) / nrow(study_a_clean)),
    sprintf("%s of %s (100%%)",
            format(nrow(study_b), big.mark = ","),
            format(nrow(study_b), big.mark = ","))
  ),
  effective_n = c(format(nrow(matched_a$full), big.mark = ","),
                  format(round(diag$ess_total[diag$method == "overlap_w"]),
                         big.mark = ",")),
  balance_achieved = c("All |SMD| <= 0.10, after searching the caliper",
                       "Exact, by construction"),
  tuning_required = c("Yes — caliper", "None")
) |> kable()
```

Weighting keeps every subject, but the effective sample size is what it actually spends,
and that number is lower than the headcount. Matching keeps fewer subjects but each one
counts once. The two are closer in real precision than the raw counts suggest — the
argument for weighting is not that it is free, it is that balance is exact and there is no
tuning parameter to get wrong.

## Saving for the outcome models

```{r}
#| label: save
saveRDS(study_b, file.path(dirname(PATHS$study_b), "study_b_weighted.rds"))
```

## What to carry forward

- Overlap weights balance exactly and are bounded, so no subject can dominate. They also
  change the estimand to the population with genuine equipoise, which has to be stated.
- Report the effective sample size. It is what the weighted analysis actually costs, and
  a headcount hides it.
- Extreme-weight thresholds are method-specific. One cutoff across methods on different
  scales is not a diagnostic.
- Balance is judged on SMD, not on p-values, in the weighted table exactly as in the
  unweighted one.

Next: [the competing-risks models](05-competing-risks.qmd) that use these weights.
