---
title: "Competing risks"
subtitle: "Problem 3 — the Fine-Gray model, and the sensitivity analyses around it"
---

```{r}
#| label: setup
source("R/setup.R")
library(survival)
library(ggplot2)
library(knitr)

study_b <- require_stage_input(
  file.path(dirname(PATHS$study_b), "study_b_weighted.rds"),
  produced_by = "notebooks/04-overlap-weighting.qmd"
)
study_b <- require_schema(study_b, schema_for("study_b_weighted"),
                          stage = "Study B competing risks")
```

## The problem

Death competes with the outcome of interest. A subject who dies can never go on to have
the event, so they are not "still at risk" in any meaningful sense — but a standard
survival analysis treats them exactly as it treats someone who moved away or reached the
end of the study. Censoring a death implicitly assumes the subject would have gone on to
experience the event at the same rate as everyone still under observation, which for a
dead subject is false.

The consequence is a cumulative incidence that is too high, and it is worst precisely
where deaths are most common — in older and sicker subgroups, and in comparisons where
the two arms have different mortality.

```{r}
#| label: prepare
analysis <- prepare_outcome_data(study_b)

status_tbl <- as.data.frame(table(analysis$cr_status))
names(status_tbl) <- c("status", "n")
status_tbl$percent <- round(100 * status_tbl$n / nrow(analysis), 2)
kable(status_tbl)
```

Competing deaths outnumber events by roughly
`r round(sum(analysis$cr_status == "death_before_event") / sum(analysis$cr_status == "event"), 1)`
to one. This is not a setting where the competing risk can be waved away.

## What this analysis is for

```{r}
#| label: contract
#| output: asis
contract_cr <- analysis_contract(
  question = paste(
    "In patients starting", STUDY_B_GROUP_LEVELS[1], "or", STUDY_B_GROUP_LEVELS[2],
    "-- does", STUDY_B_GROUP_LEVELS[2], "change the probability of experiencing",
    "the event within", STUDY_B_HORIZON_YEARS,
    "years, given that death prevents it from ever happening?"
  ),
  estimand = paste(
    "Subdistribution hazard ratio -- the quantity that maps onto a contrast in",
    "cumulative incidence, rather than a rate among survivors."
  ),
  population = paste(
    "The overlap-weighted population: subjects who could plausibly have",
    "received either treatment. Not the full source cohort."
  ),
  estimator = paste(
    "Fine-Gray model, overlap-weighted, with id = record_id and robust = TRUE",
    "for the repeated rows the expansion creates."
  ),
  causal = TRUE,
  requires = c("balance", "positivity", "events", "competing_risk")
)

treat_b <- as.integer(analysis$exposure_group == STUDY_B_GROUP_LEVELS[2])

# The balance gate is scoped to the covariates the design commits to
# balancing: those in the propensity model. See the note below on the one
# covariate deliberately left out of it.
status_cr <- validate_contract(contract_cr, evidence = list(
  balance      = balance_table(analysis, GROUP_VAR, PS_FORMULA_VARS,
                               weights = analysis$overlap_w),
  ps           = analysis$ps,
  treat        = treat_b,
  n_events     = sum(analysis$cs_event),
  n_covariates = 1,
  status       = analysis$cr_status
))

contract_block(contract_cr, status_cr)
```

The question names the competing event explicitly, and that is what forces the choice of
estimand. "Does the drug change the probability of the event" is a question about
cumulative incidence, and the subdistribution hazard is the quantity that maps onto it. A
cause-specific hazard ratio answers a different question — the rate among those still
alive — and reporting one while claiming the other is the substitution this notebook
exists to avoid.

The `competing_risk` gate is not decoration: it fails if the competing event is absent
from the data, which is exactly the situation in which someone would reach for a standard
survival model without noticing the question had changed underneath them.

### What the balance gate does and does not cover

The gate above is scoped to the covariates in the propensity model. That scope is a
decision, and writing it down is the point of having a contract at all — so here is what
it excludes and why.

```{r}
#| label: mediator-balance
med_unw <- balance_table(analysis, GROUP_VAR, BASELINE_PREDICTORS)
med_wt  <- balance_table(analysis, GROUP_VAR, BASELINE_PREDICTORS,
                         weights = analysis$overlap_w)
pick <- function(tbl) tbl$abs_smd[tbl$covariate == CLINICAL_CONTINUOUS_VAR]

data.frame(
  covariate_set = c("In the propensity model", "The excluded mediator"),
  n = c(length(PS_FORMULA_VARS), 1),
  unweighted_max_smd = round(c(
    max(balance_table(analysis, GROUP_VAR, PS_FORMULA_VARS)$abs_smd),
    pick(med_unw)), 3),
  weighted_max_smd = round(c(
    max(balance_table(analysis, GROUP_VAR, PS_FORMULA_VARS,
                      weights = analysis$overlap_w)$abs_smd),
    pick(med_wt)), 3)
) |> kable(row.names = FALSE,
           col.names = c("Covariate set", "N", "Unweighted max |SMD|",
                         "Weighted max |SMD|"))
```

`clinical_continuous_1` sits on the causal pathway, so it is excluded from the propensity
model on purpose: conditioning on it would adjust away part of the effect being estimated.
Overlap weighting balances the covariates it is given **exactly**, and gives no guarantee
about anything else. The mediator improves from
`r round(pick(med_unw), 3)` to `r round(pick(med_wt), 3)` as a side effect of balancing its
correlates, and stays above the 0.10 threshold.

That residual imbalance is expected and is not something to fix. Adjusting it away would
answer a different question — the effect *not* operating through this variable — which is
not what was asked.

It is worth contrasting with [Study A](01-balance-and-psm.qmd), where the same mediator was
excluded from the propensity model but still came in under threshold after matching.
Matching pairs subjects who are close on the whole covariate profile, so a variable
correlated with the matching set tends to come along; weighting has no such incidental
effect. Same covariate, same exclusion, different designs, different outcomes — which is
why the gate names the scope it checks instead of implying it covers everything.

## The event definition comes first

Before any model, the classification rule has to be settled, because everything
downstream inherits it. There is one function that decides it, so there is one place to
look.

```{r}
#| label: tie-rule-code
#| eval: false
# From R/simulate.R
derive_cr_status <- function(event_date, death_month_date,
                             tie_rule = c("event_wins", "death_wins")) {
  before <- if (tie_rule == "event_wins") {
    event_date <= death_month_date   # an event on the death date counts as an event
  } else {
    event_date <  death_month_date   # ... or as a competing death
  }
  ...
}
```

The reason a tie rule is needed at all is a property of the data rather than of the
statistics. Death is recorded at **month precision** — every death date has a day
component of 01 — while the event of interest carries true day precision. So an event
recorded on the first of the month in which someone died lands on exactly the same date
as their death, and the comparison has to say which one wins.

```{r}
#| label: tie-analysis
strict <- prepare_outcome_data(study_b, tie_rule = "death_wins")

n_ties <- sum(analysis$cr_status != strict$cr_status)

raw_both <- !is.na(study_b$event_date) & !is.na(study_b$death_month_date)
n_inverted <- sum(study_b$.true_event_first == 1 & raw_both &
                    study_b$event_date > study_b$death_month_date)

data.frame(
  quantity = c("Subjects whose status changes under the strict rule",
               "Subjects with event and death in the same month",
               "Subjects whose true ordering is inverted by month rounding"),
  n = c(n_ties,
        sum(raw_both & format(study_b$event_date, "%Y-%m") ==
              format(study_b$death_month_date, "%Y-%m")),
        n_inverted)
) |> kable()
```

The honest result here is that the tie rule moves very few subjects — `r n_ties` in this
cohort. That is worth reporting rather than dressing up, and it is not an argument for
skipping the check: the number could not be known without computing it, and a
specification that leaves the rule implicit is one where two analysts get different
answers and neither can say why.

The more consequential finding is the third row. Month-precision rounding pushes the
recorded death date *earlier*, so `r n_inverted` subjects whose event genuinely came first
have a recorded event date that falls after their recorded death date. Those subjects are
classified as competing deaths under **both** rules. The tie rule cannot fix this, because
it is not a tie — it is a rounding artifact that inverts the true ordering, and the only
real remedy is a death date with day precision.

```{r}
#| label: tie-invariance
# Follow-up time is bounded by the minimum over both dates symmetrically, so
# the tie rule can change the label attached to a subject but never the time.
stopifnot(identical(round(analysis$fu_years, 10), round(strict$fu_years, 10)))
```

That the tie rule cannot change follow-up time is worth asserting rather than assuming.
It means the rule affects only the classification, and no downstream time-to-event
quantity moves underneath it.

## The Fine-Gray model

Two hazards can be estimated here, and they answer different questions.

The **cause-specific** hazard asks: among those still alive and event-free, what is the
rate of the event? It is the right quantity for questions about biological mechanism.

The **subdistribution** hazard, which the Fine-Gray model estimates, keeps subjects who
died in the risk set with a declining weight. That sounds strange as a description of who
is at risk, and it is — but it is what makes the coefficient map directly onto the
cumulative incidence function. It is the right quantity for questions about what
proportion of patients will actually experience the event.

```{r}
#| label: finegray
# ow and record_id appear on the right-hand side purely so finegray() carries
# them into the expanded dataset. Neither is a predictor.
fg_data <- finegray(
  Surv(fu_years, cr_status) ~ exposure_group + overlap_w + record_id,
  data = analysis, etype = "event"
)

# The censoring weight from the Fine-Gray expansion multiplies the overlap
# weight: one handles the competing risk, the other handles confounding.
fg_data$analysis_weight <- fg_data$fgwt * fg_data$overlap_w

data.frame(
  quantity = c("Subjects before expansion", "Rows after expansion",
               "Row inflation", "Distinct subjects preserved"),
  value = c(format(nrow(analysis), big.mark = ","),
            format(nrow(fg_data), big.mark = ","),
            sprintf("%.1fx", nrow(fg_data) / nrow(analysis)),
            format(length(unique(fg_data$record_id)), big.mark = ","))
) |> kable()

stopifnot(
  length(unique(fg_data$record_id)) == nrow(analysis),
  all(abs(fg_data$analysis_weight - fg_data$fgwt * fg_data$overlap_w) < 1e-12)
)
```

```{r}
#| label: fg-fit
fit_fg <- coxph(
  Surv(fgstart, fgstop, fgstatus) ~ exposure_group,
  data = fg_data, weights = analysis_weight, id = record_id, robust = TRUE
)

fit_cs <- coxph(
  Surv(fu_years, cs_event) ~ exposure_group,
  data = analysis, weights = overlap_w, id = record_id, robust = TRUE
)

primary <- rbind(
  model_result_row(fit_fg, term = 1L, label = "Primary",
                   model = "Fine-Gray (subdistribution)"),
  model_result_row(fit_cs, term = 1L, label = "Primary",
                   model = "Cause-specific Cox")
)
primary <- add_display(primary)
primary$true_cs_hr <- round(exp(TRUE_B_EVENT[["treat"]]), 3)

primary[, c("model", "hr_ci", "p", "true_cs_hr")] |>
  kable(col.names = c("Model", "HR (95% CI)", "p", "True cause-specific HR"),
        row.names = FALSE)

save_table(primary, "competing_risks_primary")
```

### How the two estimates relate

The gap between a subdistribution and a cause-specific hazard ratio is driven by what
treatment does to the **competing** hazard. So that model has to be fitted too.

```{r}
#| label: death-model
analysis$death_event <- as.integer(analysis$cr_status == "death_before_event")

fit_death <- coxph(
  Surv(fu_years, death_event) ~ exposure_group,
  data = analysis, weights = overlap_w, id = record_id, robust = TRUE
)

three <- rbind(
  primary[, setdiff(names(primary), c("hr_ci", "p", "true_cs_hr"))],
  model_result_row(fit_death, term = 1L, label = "Primary",
                   model = "Competing death, cause-specific")
)
three <- add_display(three)
three$true_hr <- c(round(exp(TRUE_B_EVENT[["treat"]]), 3),
                   round(exp(TRUE_B_EVENT[["treat"]]), 3),
                   round(exp(TRUE_B_DEATH[["treat"]]), 3))

three[, c("model", "hr_ci", "p", "true_hr")] |>
  kable(col.names = c("Model", "HR (95% CI)", "p", "True HR"),
        row.names = FALSE)
```

The two event estimates are very close, and that is the expected result here: treatment
was generated with essentially no effect on mortality, so both arms accumulate competing
deaths at a similar rate, those deaths hold down the cumulative incidence equally in both
arms, and nothing pulls the two estimands apart. The place where they diverge sharply is a
treatment that changes mortality substantially — which is exactly when reporting the wrong
one does the most damage.

```{r}
#| label: estimand-comparison
sub_hr <- primary$estimate[primary$model == "Fine-Gray (subdistribution)"]
cs_hr  <- primary$estimate[primary$model == "Cause-specific Cox"]

stopifnot(
  # Both recover a protective effect...
  sub_hr < 1, cs_hr < 1,
  # ...the cause-specific model recovers the truth it targets...
  abs(log(cs_hr) - TRUE_B_EVENT[["treat"]]) < 0.10,
  # ...and with a near-null competing hazard the two estimands nearly coincide.
  abs(log(sub_hr) - log(cs_hr)) < 0.05
)
```

One detail in that table is worth not glossing over. The estimated competing-death hazard
ratio is `r round(three$estimate[3], 3)`, above 1, while the value used to generate the
data was `r round(exp(TRUE_B_DEATH[["treat"]]), 3)` — essentially null. This is not a
weighting failure: the same weights recover the event hazard ratio accurately. It is a
selection effect. Treatment reduces the event, so treated subjects are censored by the
event less often and remain at risk for death longer, and the subjects who avoid the event
are not a random subset — the shared frailty in the data generator makes the ones who
remain systematically frailer.

That is a general hazard of reading cause-specific models of the *competing* event as if
they were causal: the risk set has been shaped by the treatment's effect on the other
cause. It is also why the competing-death model belongs in the report as a descriptive
companion rather than as a second treatment-effect estimate.

Both event estimands are correct answers to different questions. Reporting one and
describing it as the other is the error to avoid — a paper that says "risk of the event"
while presenting a cause-specific hazard ratio has made exactly that substitution.

## Why the robust variance

Every model here is fitted with `id = record_id, robust = TRUE`, and every interval and
p-value is computed from the sandwich variance stored in `fit$var` rather than the
model-based variance in `fit$naive.var`. Three reasons, all of which apply
simultaneously.

**Overlap weights are analytic weights, not frequency weights.** `coxph` treats the
`weights=` argument as though it counted observations — a weight of 3 means three
identical subjects. That is not what an overlap weight means: it expresses how much a
subject contributes to a target population, not how many of them there are. The
model-based variance inherits the frequency-weight interpretation and understates the
uncertainty accordingly.

**The Fine-Gray expansion produces many rows per subject.** The
`r sprintf("%.1fx", nrow(fg_data) / nrow(analysis))` inflation above is one subject
appearing repeatedly across time intervals. Those rows are not independent observations,
and treating them as such is precisely the error that `id=` plus a sandwich estimator
exists to correct.

**Consistency across the report.** Mixing estimators between tables produces numbers that
cannot be compared with each other, and a reader has no way to tell which is which.

The practical consequence is that `summary(fit)` prints both a `se(coef)` and a
`robust se`, and reading the wrong column produces a different confidence interval and a
different p-value from the *same fitted model* and the *same coefficient*. This project
avoids the ambiguity structurally: `model_result_row()` takes the variance source as an
argument, computes the Wald statistic, interval and p-value from that one source, and
records which was used in a `se_source` column that travels with the estimate.

```{r}
#| label: se-provenance
data.frame(
  model = primary$model,
  se_log_hr = round(primary$se_log, 4),
  se_source = primary$se_source
) |> kable(row.names = FALSE)
```

## Sensitivity analyses

A primary estimate on its own says very little. What makes it credible is that the
analyses around it — different horizons, a different weighting scheme, a different
confounding-control strategy, a different handling of early events — point the same way.
If one of them disagreed, that would be the finding.

```{r}
#| label: sensitivity-helpers
fit_fg_at <- function(data, label, weight_var = "overlap_w", horizon = NULL) {
  d <- data
  if (!is.null(horizon)) {
    d$cr_status <- as.character(d$cr_status)
    d$cr_status[d$fu_years > horizon] <- "censored"
    d$cr_status <- factor(d$cr_status,
                          levels = c("censored", "event", "death_before_event"))
    d$fu_years <- pmin(d$fu_years, horizon)
  }
  d$.w <- d[[weight_var]]

  fg <- finegray(Surv(fu_years, cr_status) ~ exposure_group + .w + record_id,
                 data = d, etype = "event")
  fg$analysis_weight <- fg$fgwt * fg$.w

  fit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ exposure_group,
               data = fg, weights = analysis_weight, id = record_id,
               robust = TRUE)
  model_result_row(fit, term = 1L, label = label, model = "Fine-Gray")
}
```

```{r}
#| label: landmark
# 180-day landmark. Everyone with follow-up at or below the landmark is
# excluded and the clock restarts for the rest. Because fu_years is already the
# time to the first of event, death or loss to follow-up, the single condition
# removes early events, early deaths and early dropout together.
lag_years <- STUDY_B_LANDMARK_DAYS / DAYS_PER_YEAR
keep <- analysis$fu_years > lag_years
landmark <- analysis[keep, ]
landmark$fu_years <- landmark$fu_years - lag_years

data.frame(
  quantity = c("Excluded before the landmark", "Retained", "Events retained"),
  n = c(sum(!keep), sum(keep), sum(landmark$cr_status == "event"))
) |> kable()

stopifnot(all(landmark$fu_years > 0),
          nrow(landmark) + sum(!keep) == nrow(analysis))
```

The team called this a "180-day lag", but methodologically it is a landmark analysis: the
cohort is restricted to survivors at day 180 and follow-up is measured from there. It
addresses reverse causation — events occurring immediately after the index date are
unlikely to have been caused by a treatment started that day — at the cost of excluding
everyone with short follow-up.

```{r}
#| label: sensitivity-run
sens <- rbind(
  primary[primary$model == "Fine-Gray (subdistribution)",
          setdiff(names(primary), c("hr_ci", "p", "true_cs_hr"))] |>
    transform(analysis = "Primary (3-year horizon, overlap weights)"),
  fit_fg_at(analysis, "1-year horizon", horizon = 1),
  fit_fg_at(analysis, "2-year horizon", horizon = 2),
  fit_fg_at(analysis, "Stabilized IPTW instead of overlap", weight_var = "sw_iptw"),
  fit_fg_at(landmark, "180-day landmark")
)
```

```{r}
#| label: sensitivity-adjusted
# A regression-adjusted, unweighted alternative: a different strategy for
# confounding control rather than a different weight. This is the expensive fit
# -- the covariates are carried through the expansion.
adj_terms <- paste(PS_FORMULA_VARS, collapse = " + ")
fg_adj <- finegray(
  as.formula(paste("Surv(fu_years, cr_status) ~ exposure_group +",
                   adj_terms, "+ record_id")),
  data = analysis, etype = "event"
)
fit_adj <- coxph(
  as.formula(paste("Surv(fgstart, fgstop, fgstatus) ~ exposure_group +", adj_terms)),
  data = fg_adj, weights = fgwt, id = record_id, robust = TRUE
)

term_idx <- grep("^exposure_group", names(coef(fit_adj)))[1]
sens <- rbind(sens, model_result_row(
  fit_adj, term = term_idx,
  label = "Multivariable-adjusted, unweighted", model = "Fine-Gray"
))
```

```{r}
#| label: sensitivity-table
sens_display <- add_display(sens)
sens_display[, c("analysis", "hr_ci", "p", "se_source")] |>
  kable(col.names = c("Analysis", "sHR (95% CI)", "p", "Variance"),
        row.names = FALSE)

save_table(sens, "competing_risks_sensitivity")
```

```{r}
#| label: forest
#| fig-height: 5
sens_display$analysis <- factor(sens_display$analysis,
                                levels = rev(sens_display$analysis))

ggplot(sens_display, aes(x = estimate, y = analysis)) +
  geom_vline(xintercept = 1, colour = "grey60") +
  geom_pointrange(aes(xmin = ci_lower, xmax = ci_upper), size = 0.4) +
  scale_x_log10() +
  labs(x = "Subdistribution hazard ratio (95% CI, log scale)", y = NULL,
       title = "The primary estimate and the analyses that have to agree with it") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: report-consistency
#| output: asis
# Agreement between the sensitivity analyses and the primary estimate is a
# property of the data, not of the code. It is reported, never asserted: a
# divergence is the most informative thing this analysis could surface, and a
# pipeline that halted on it would suppress exactly the result worth
# investigating. Contrast with the contract gates above, which do stop --
# those describe whether the analysis is capable of answering its question.
primary_est <- sens$estimate[sens$analysis == "Primary (3-year horizon, overlap weights)"]
agreement <- report_agreement(sens$estimate, sens$analysis, reference = primary_est)
```

```{r}
#| label: agreement-table
kable(agreement, row.names = FALSE,
      col.names = c("Analysis", "sHR", "Side of null", "Agrees with primary"))
```

The multivariable-adjusted model is the one worth reading separately: it controls
confounding by regression rather than by weighting, and it estimates a conditional rather
than a marginal effect, so it is not expected to reproduce the primary estimate exactly —
only to agree with it in direction and rough magnitude.

## Proportional hazards

`cox.zph()` is not used on the Fine-Gray fit. Its risk set is artificial by construction —
subjects who have died remain in it with declining weights — so the scaled Schoenfeld
residuals it produces have no interpretation as a diagnostic for that model. Instead
follow-up is split into intervals and a treatment-by-period interaction is tested
jointly, on the robust covariance.

```{r}
#| label: ph-test
ph <- rbind(
  ph_interval_test(analysis, "fu_years", "cs_event", GROUP_VAR,
                   weight_var = "overlap_w", id_var = "record_id",
                   cuts = seq_len(STUDY_B_HORIZON_YEARS - 1),
                   label = "Cause-specific Cox, overlap weighted")
)
ph$p_value <- fmt_p(ph$p_value)
kable(ph, row.names = FALSE)
save_table(ph, "competing_risks_ph")
```

The data were generated from a proportional hazards model, so the assumption genuinely
holds here and the test agreeing with that is a check on the test rather than a discovery.
The wording of the conclusion is deliberately weak: a non-significant result is an absence
of evidence against proportionality, not evidence for it.

## Cumulative incidence

```{r}
#| label: cif
#| fig-height: 5
cif <- survfit(Surv(fu_years, cr_status) ~ exposure_group,
               data = analysis, weights = overlap_w, id = record_id)

cif_df <- data.frame(
  time = cif$time,
  cif = cif$pstate[, which(cif$states == "event")],
  arm = rep(STUDY_B_GROUP_LEVELS, cif$strata)
)

ggplot(cif_df, aes(x = time, y = cif, colour = arm)) +
  geom_step(linewidth = 0.8) +
  scale_colour_manual(values = c("#2c3e50", "#c0392b")) +
  labs(x = "Years since index", y = "Cumulative incidence of the event",
       colour = NULL,
       title = "Overlap-weighted cumulative incidence, accounting for competing death") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

This curve is what the Fine-Gray coefficient describes. A Kaplan-Meier plot of the same
data would rise higher in both arms, because it treats every competing death as though the
subject remained at risk.

## What to carry forward

- Settle the event definition before fitting anything, in one function, and verify how
  much the choice actually moves. Here the tie rule moves `r n_ties` subjects, and the
  month-precision rounding underneath it moves more.
- Cause-specific and subdistribution hazards answer different questions. Report both, and
  name which one any given number is.
- Analytic weights and repeated rows per subject both require the robust sandwich
  variance. Read `fit$var`, not `fit$naive.var`, and record which was used alongside the
  estimate so the two can never be confused downstream.
- A primary estimate is only as good as the sensitivity analyses that agree with it.
- `cox.zph()` has no valid interpretation for a Fine-Gray fit.
